An algorithm for finding the global extremum of a partially defined function\thanks{This work was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-08-29 2023/MMST/samplepaper.tex
fix terminology

This paper presents the results in a new research direction related to the use of imputation\footnote{The term ``imputation'' in the paper is used following the established terminology in the field of machine learning related to the restoring of the missing data.} (restoring) of the objective function values in the course of the global search algorithm operation when solving the problems with partially defined objective functions. The approach proposed is based on an adoption of the computation rules of the global search algorithm. The efficiency of the approach proposed has been demonstrated numerically.

@@

  \item For each interval $(x_{i-1}, x_i), 1 \leq i \leq k+1$, the computing of the characteristic $R(i)$ is performed according to the following computing formulae.

        \paragraph{Case 1} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the classic rules of global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 2}   For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 3} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0}
  \end{cases}
\end{equation}
        \paragraph{Case 4} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, v(x_{i-1})=v(x_i)=0
\end{equation}
        \paragraph{Case 5} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the computable interval with the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

\subsection{Results of Numerical Experiments}
The numerical experiments were carried out using a computer with Intel® Core™ i7-10750H CPU @ 2.60GHz 2.59 GHz using the developed software package implementing the modification described in Section 3. To generate a series of problems, GKLS generator described in \cite{Gaviano2003} was used. It allows generating the multiextremal oprimization problems with the properties known in advance. These problems were appended with the non-computable regions of various kinds.

Each series of experiments consisted of 100 GKLS problems with everywhere computable objective function (the classic GKLS problems), 100 problems with the non-computabilities at the boundaries of the search domains, and 100 problems with non-computability regions defined randomly.

In the first series of experiments, the correctness of the modification operation was checked and the features of executing the trials in the non-computability regions were revealed. The problems of the dimensionality $N=2$ were used in the experiment. The reliability parameter was set to $r=5.5$, the precision $\varepsilon=0.01$. The results of experiments are presented in Tab.~\ref{tab1}.

\begin{figure}
\includegraphics[width=\textwidth]{fig6.png}
\caption{Examples of problems \#\# 19, 25, and 35 from the first series of experiments} \label{fig6}
\end{figure}

Fig.~\ref{fig6} presents the examples of the modification operation with several problems of the series. The trials inside the computable regions are denoted by the blue points in the figures, the global minima of the problems – by the red points, and the points of trials performed inside the non-computability regions – by black. Also, the non-computability regions are filled with grey for clarity.

Since the modification works with the classic GKLS problem series according to the base global search algorithm, one can compare it with the classic GSA. Thus, the general behavior of GSA was preserved for the modification, and the same regions of trials’ densification near the local minima were observed that can be explained by the realization of the classical computation rules (\ref{eq21}) in most cases. One can see in Tab.~\ref{tab1} that the number of trials executed inside the non-computable regions is small. Total number of trials for the classic problems and for the problems with non-computable regions are comparable, the method doesn’t perform excess iterations in the non-promising regions due to the computing formulae utilizing the information on the nearest computable points.

\begin{table}
\caption{Comparison of the efficiencies of the algorithm modification with the classic version of the global search algorithm on GKLS problems}\label{tab1}
\begin{tabular}{|l|c|c|c|}
\hline
 &  classical  &  round district  &  random district  \\
\hline
Average number of iterations & 781,66	& 854,91 & 839,03 \\
Average number of non-computability & - & 185,79 & 42,31 \\
Number of solved problems & 100 / 100 & 100 / 100 & 100 / 100\\
\hline
\end{tabular}
\end{table}

@@

100\% of the problems were solved. However, for the series with the boundary non-computability region, the value of parameter $r=6.0$ was required to solve one problem, for the series with the random regions – soving of two problems with given value of parameter $r$.

In the second series of experiments, the problem series of various dimensionalities were solved. In the experiments with the problems of the dimensionality $N=4$, the reliability parameter was set to $r=4.5$, the precision $\varepsilon=0.02$. The global minimum was considered to be found if the algorithm generates a trial point inside the $\varepsilon$-nearness of the global minimum. The results of the experiments are presented in Tab.~\ref{tab2}.

\begin{table}
\caption{Problem series of various dimensionalities}\label{tab2}
\begin{tabular}{|l|l|c|c|c|}
\hline
 N &  Task type & Average num. of iters & Average num. of non-comput. & Solved problems \\
\hline
$3$ & classical & 12251,11 & - & 100 / 100 \\
$ $ & round & 12201,78 & 4763,57 & 100 / 100 \\
$ $ & random & 12210,71 & 63,39 & 100 / 100 \\
\hline
$4$ & classical & 25144,46 & - & 100 / 100 \\
$ $ & round & 29840,32 & 15004,37  & 100 / 100 \\
$ $ & random & 25287,49 & 26,43 & 100 / 100 \\
\hline
$5$ & classical & 56725,95 & - & 100 / 100 \\
$ $ & round & 47840,09 & 24109,19 & 100 / 100\\
$ $ & random & 57874,76 & 82,57 & 100 / 100\\
\hline
\end{tabular}
\end{table}

According to the results of the experiments, the method resolves successfully the occasional falls into the non-computability regions scattered inside the search domain randomly whereas in the case of large boundary regions it tries the execute a complete investigation without executing too many excess trials.

\subsection{Conclusions}
In the present work, the problem of finding the global minimum of a computation-costly "black box"\ function was considered. As compared to the traditional formulation used in the global optimization, the problem considered has an important difference: the objective function may be undefined in some regions of the search domain (in particular, in one point or in several ones). The appearing of such a problem statement is caused by a frequent problem of numerical instability of the models investigated in the applied problems.

The main problem is the impossibility to identify the non-computability regions in advance in most cases. In fact, the region of feasible combinations of parameters is undefined a priori. This phenomenon can be interpreted as the presence of additional hidden constraints in the problem.

The problem investigated in the present work echoes to problem of missing data existing in Data Science for a long time. As a solution, the experts in the field have proposed various approaches to the imputation (restoring) of the missing values. These ideas have found their reflection in the development of the new method.

The applicability of already existing approaches appended with various variants to account for the Lipschitz condition was investigated. As a result, a universal efficient algorithm for solving the problems of this class has been developed. The algorithm constructed is a development of one of efficient deterministic method of solving the Lipschitz global optimization problems, namely information-statistical global search algorithm. Within the framework of this modification, the computation rules of the algorithm were adopted with introduction of the data imputation algorithm.

The numerical experiments were carried out using series of multiextremal GKLS problems of the dimensionalities $N = 2,3,4,5$. The results have demonstrated the reliability and efficiency of the approach proposed. The modified algorithm handles successfully the problems, which the non-computability regions are scattered randomly within the search domain. In the case of large boundary non-computability regions, the method solves the problems efficiently as well performing a complete investigation of the non-computability regions also.

>>>>

This paper presents the results in a new research direction related to the use of imputation\footnote{The term ``imputation'' in the paper is used following the established terminology in the field of machine learning related to the restoring of the missing data.} (restoring) of the objective function values in the course of the global search algorithm operation when solving the problems with partially defined objective functions. The approach proposed is based on an adaption of the computation rules of the global search algorithm. The efficiency of the approach proposed has been demonstrated numerically.

@@

  \item For each interval $(x_{i-1}, x_i), 1 \leq i \leq k+1$, the computing of the characteristic $R(i)$ is performed according to the following formulae.

        \paragraph{Case 1.} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the original rules of the global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1,}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1,}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1.}
  \end{cases}
\end{equation}
        \paragraph{Case 2.}   For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0,}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1.}
  \end{cases}
\end{equation}
        \paragraph{Case 3.} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1,}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0.}
  \end{cases}
\end{equation}
        \paragraph{Case 4.} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, \; v(x_{i-1})=v(x_i)=0.
\end{equation}
        \paragraph{Case 5.} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the interval with computable boundary points that has the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

\section{Results of Numerical Experiments}

The numerical experiments were carried out using a computer with Intel® Core™ i7-10750H CPU @ 2.60GHz using the developed software package implementing the modification described in Section 3. To generate a series of problems, GKLS generator described in \cite{Gaviano2003} was used. It allows generating the multiextremal optimization problems with the properties known in advance. These problems were appended with the non-computable regions of various kinds.

Each series of experiments consisted of 100 GKLS problems with everywhere computable objective function (the original GKLS problems), 100 problems with the non-computabilities at the boundaries of the search domains, and 100 problems with non-computability regions defined randomly.

In the first series of experiments, the correctness of the modification operation was checked and the features of executing the trials in the non-computability regions were revealed. The problems of the dimensionality $N=2$ were used in the experiment. The reliability parameter was set to $r=5.5$, the accuracy was $\varepsilon=0.01$. The results of experiments are presented in Table~\ref{tab1}.

\begin{table}[h]
\caption{Comparison of the original and modified GSA}\label{tab1}
\begin{tabular}{|l|c|c|c|}
\hline
 &  classical  &  round district  &  random district  \\
\hline
Average number of trials & 781,66	& 854,91 & 839,03 \\
Average number of trials in non-computability regions & - & 185,79 & 42,31 \\
Number of solved problems & 100 / 100 & 100 / 100 & 100 / 100\\
\hline
\end{tabular}
\end{table}

\begin{figure}[h]
\includegraphics[width=\textwidth]{fig6.png}
\caption{Examples of problems \#\# 19, 25, and 35 from the first series of experiments} \label{fig6}
\end{figure}

Fig.~\ref{fig6} presents the examples of solving several problems with the modified GSA. The trials inside the computable regions are marked by the blue points, the global minima of the problems are marked by the red points, and the points of trials performed inside the non-computability regions are marked by the black points. Also, the non-computability regions are filled with grey for clarity.

Since the modified algorithm works with the original GKLS problem series according to the rules of base algorithm, one can compare it with the GSA. Thus, the general behavior of GSA was preserved for the modification, and the same regions of trials' accumulation near the local minima were observed that can be explained by the realization of the  computation rules (\ref{eq21}) in most cases. One can see in Table~\ref{tab1} that the number of trials executed inside the non-computable regions is small. Total number of trials for the original problems and for the problems with partially defined objective function are comparable, the method doesn't perform excess trials in the non-promising regions due to the computing formulae utilizing the information on the nearest computable points.

@@

100\% of the problems were solved. However, for the series with the boundary non-computability region the value of parameter $r=6.0$ was required to solve one problem, and for the series with the random non-computability regions the same value of $r$ was required to solve two problems.

In the second series of experiments, the problem series of various dimensions were solved. In the experiments with the problems of the dimensions $N=3,4,5$, the reliability parameter was set to $r=4.5$, the precision $\varepsilon=0.02$. The global minimum was considered to be found if the algorithm generates a trial point inside the $\varepsilon$-nearness of the global minimum. The results of the experiments are presented in Table~\ref{tab2}.

\begin{table}
\caption{Problem series of various dimensions}\label{tab2}
\begin{tabular}{|l|l|c|c|c|}
\hline
 N &  Task type & Average num. of trials & Average num. of non-comput. & Solved problems \\
\hline
$3$ & classical & 12251.1 & - & 100 / 100 \\
$ $ & round & 12201.8 & 4763.6 & 100 / 100 \\
$ $ & random & 12210.7 & 63.4 & 100 / 100 \\
\hline
$4$ & classical & 25144.5 & - & 100 / 100 \\
$ $ & round & 29840.3 & 15004.4  & 100 / 100 \\
$ $ & random & 25287.5 & 26.4 & 100 / 100 \\
\hline
$5$ & classical & 56725.9 & - & 100 / 100 \\
$ $ & round & 47840.1 & 24109.2 & 100 / 100\\
$ $ & random & 57874.8 & 82.6 & 100 / 100\\
\hline
\end{tabular}
\end{table}

The results of the experiments show that the method resolves successfully the hits into the non-computability regions randomly scattered inside the search domain whereas in the case of large boundary regions it tries the execute a complete exploration without executing too many excess trials.

\section{Conclusions}

In the present work, the problem of finding the global minimum of a computation-costly ``black box'' function was considered. As compared to the traditional formulation used in the global optimization, the considered problem has an important difference: the objective function may be undefined in some regions of the search domain. The appearing of such a formulation is caused by a frequent problem of numerical instability of the models investigated in the applied tasks.

The main problem is the impossibility to identify the non-computability regions in advance in the most cases. In fact, the region of feasible combinations of parameters is unknown a priori. This phenomenon can be interpreted as partial computability of the objective function.

The problem investigated in the present work echoes to problem of missing data existing in Data Science for a long time. As a solution, the experts in the field have proposed various approaches to the imputation (restoring) of the missing values. %These ideas have found their reflection in the development of the new method.
The applicability of already existing approaches appended with the different variants to account for the Lipschitz condition was investigated. As a result, efficient algorithm for solving the problems of this class has been proposed. This algorithm is a modification of the deterministic method for solving the Lipschitz global optimization problems, namely information-statistical global search algorithm.
%Within the framework of this modification, the scheme of the algorithm were adopted with introduction of the data imputation rules.
As part of the modification, the computational scheme of the algorithm was adapted with the introduction of the data imputation rules.

The numerical experiments were carried out using series of multiextremal GKLS problems of the dimensions $N = 2,3,4,5$. The results have demonstrated the reliability and efficiency of the proposed approach. The modified algorithm handles successfully the problems, in which the non-computability regions are randomly scattered within the search domain. In the case of large boundary non-computability regions, the method solves the problems efficiently as well performing an exploration of the non-computability regions.
